Well-quasi-orderings on word languages

### 2024-10-18 src/amalgam.tex
new proofs

\AP
In this section, we are going to design an effective decision procedure for
\kl{well-quasi-ordering} by the \kl{infix relation}. To that end, the first
requirement is to fix a way to represent languages $L \subseteq \Sigma^*$.
Traditionally, one would use finite automata, context-free grammars, or for the
more adventurous, vector addition systems with states. However, our proof
technique will only require us to have a way to ``glue'' together runs of the
system to ``pump'' them and produce new runs: this is the usual pumping lemma
in automata theory, and Ogden's lemma for context-free grammars.
\todo{Cite the pumping lemma for VASS. Cite Ogden's lemma.}

\AP It turns out that there is a rather large family of systems for which
pumping arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anad, Schmitz,
Sch\"{u}tze, and Zetzsche \cite{ASZZ24}. Having done the heavy lifting on
\kl{bounded languages}, the rest of this section is mostly devoted to the
introduction of \kl{amalgamation systems} and collecting the necessary pumping
argument they enjoy. Using this meta-proof, we will generalize
\cref{infix-finite-automata:thm} to context-free grammars, languages recognized
by vector addition systems with state (VASS), and more.
\todo{unclear formulation}
The goal of this
section is not to introduce all of these systems, or justify their usefulness,
but to state and prove the following two theorems.

\begin{theorem}[label=infix-amalgamation:thm,restate=infix-amalgamation:thm]
    \proofref{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    \kl(language){bounded}.
\end{theorem}

\begin{theorem}[restated=infix-wqo-is-emptiness:thm,label=infix-wqo-is-emptiness:thm]
% 	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be a class of languages recognized by 
    \kl{amalgamation systems}
    and closed under
    \kl{rational transductions}. 
    Then the following are equivalent:
	\begin{enumerate}
        \item\label{wqo-infix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{infix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{wqo-prefix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{prefix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{emptiness-decidable} Emptiness is decidable for languages in $\mathcal{C}$.
	\end{enumerate}
\end{theorem}

\subsection{Amalgamation Systems}
\label{infixes-amalgamation:subsec}

Let us now formally introduce the notion of \kl{amalgamation systems}, and
recall some results from \cite{ASZZ24} that will be useful for the proof of
\cref{infix-amalgamation:thm}. The notion of \kl{amalgamation system} is
tailored to produce \emph{pumping arguments}, which is exactly what our
\cref{pumping-periods:lem} talks about. At the core of a pumping argument,
there is a notion of \emph{run}, which could for instance be a sequence of
transitions taken in a finite state automaton. Continuing on the analogy with
finite automata, there is a natural ordering between runs, i.e., a run is
smaller than another one if one can ``unroll loops" of the first run to obtain
the second one. Typical pumping arguments then rely on the fact that
\emph{minimal} runs are of finite size, and that all other runs are
obtained by ``gluing" minimal runs together. This is exactly what
\kl{amalgamation systems} are about.

@@

One can leverage the same kind of reasoning to take the opposite direction, and
notice that the notions of being a \kl{bounded language}, a \kl{regular
language}, or being included in a finite union of products of \kl{chains} all
collapse for \kl{well-quasi-ordered} and \kl{downwards closed} languages for
the \kl{infix relation}.

@@

@@

For the next result, we impose a mild restriction by requiring the language
class to be closed under \kl{rational transductions}. This property holds for
most common language classes, including all known \kl{amalgamation systems}.

\AP A class of languages $\mathcal{C}$ is \intro{recognized by amalgamation
systems} when, for every language $L \in \mathcal{C}$, there exists an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$ such that $L$ is the
language of this system.

\AP A \intro{rational transduction} is a function $f \colon \Sigma^* \to
\Gamma^*$ that can be defined by a finite automaton with output, or
equivalently by a finite set of rules of the form $u \to v$ where $u \in
\Sigma^*$ and $v \in \Gamma^*$. We write $f(L) = \setof{f(w)}{w \in L}$ for the
image of a language $L$ by a rational transduction $f$. A class of languages
$\mathcal{C}$ is \intro{closed under rational transductions} when, for every
language $L \in \mathcal{C}$, and every rational transduction $f \colon
\Sigma^* \to \Gamma^*$, the language $f(L)$ is in $\mathcal{C}$.\todo{check these definitions ...}

@@

    Let $R = \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in \Sigma^{\leq m
    \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup
    \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of
    the bounds $n$, $m$, and $N_0$, this language is regular. As $\mathcal{C}$
    is \kl{closed under rational transductions}, we can therefore reduce the
    inclusion to emptiness of $L \cap \overline{R}$.\todo{explain that $L
    \mapsto L \cap R$ is a rational  transduction?} However, we need to find
    these bounds first.

    To determine values for $n$ and $m$, we first test if $L$ is bounded. As
    the algorithm in \cite[Section 4.2]{ASZZ24} is effective, this yields words
    $w_1, \ldots w_n$ such that $L \subseteq w_1^* \cdots w_n^*$ and therefore
    yields also the bounds in questions.

@@

    \cref{wqo-infix-decidable} $\Rightarrow$ \cref{wqo-prefix-decidable}. We
    just consider the transduction

    \cref{wqo-prefix-decidable} $\Rightarrow$
    \cref{emptiness-decidable}. We consider the transduction $\mathcal T
    = \Sigma^* \times \{a, b\}$.\todo{Define this transduction explicitly} Then
    for any language $L \in \mathcal C$, $\mathcal TL$ is well-quasi-ordered by
    infix if and only if $L$ is empty.
\end{proofof}

>>>>

\AP In this section, we are going to design an effective decision procedure for
\kl{well-quasi-ordering} by the \kl{infix relation}. To that end, the first
requirement is to fix a way to represent languages $L \subseteq \Sigma^*$.
Traditionally, one would use finite automata, context-free grammars, or for the
more adventurous, vector addition systems with states. However, our proof
technique will only require us to have a way to ``glue'' together runs of the
system to ``pump'' them and produce new runs: this is the usual pumping lemma
in automata theory, and Ogden's lemma for context-free grammars \cite{OGDEN68}.
It turns out that there is a rather large family of systems for which pumping
arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anad, Schmitz,
Schütze, and Zetzsche \cite{ASZZ24}.

\AP Our first result, of theoretical nature, is that \kl{amalgamation systems}
cannot define \kl{well-quasi-ordered} languages that are not
\kl(language){bounded}. This implies that all the results of
\cref{infixes-regular:sec}, and in
particular \cref{bounded-language:thm}, can safely be transferred to
\kl{amalgamation systems}.

\begin{theorem}[label=infix-amalgamation:thm,restate=infix-amalgamation:thm]
    \proofref{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    If $L$ is \kl{well-quasi-ordered} by the \kl{infix relation} then $L$ is
    \kl(language){bounded}.
\end{theorem}

\AP Our second focus is of practical nature: we want to give a decision
procedure for being \kl{well-quasi-ordered}. This will require us to introduce
\emph{effectiveness assumptions} on the \kl{amalgamation systems}. While most
of them will be innocuous, an important consequence is that we have to consider
\emph{classes of languages} rather than individual ones. Such classes will be
called \kl{effective amalgamative classes} (\kcref{effective amalgamative
classes}). In the following theorem, we prove that under such assumptions,
testing \kl{well-quasi-ordering} is inter-reducible to testing whether a
language of the class is empty.

\begin{theorem}[restated=infix-wqo-is-emptiness:thm,label=infix-wqo-is-emptiness:thm]
% 	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be an \kl{effective amalgamative class} of languages.
    Then the following are equivalent:
	\begin{enumerate}
        \item\label{wqo-infix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{infix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{wqo-prefix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{prefix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{emptiness-decidable} Emptiness is decidable for languages in $\mathcal{C}$.
	\end{enumerate}
\end{theorem}

\AP We say that a \kl{strongly effective amalgamative class} is an
\kl{effective amalgamative class} for which the emptiness problem is decidable.
Let us immediately remark that the class of \kl{regular languages} is
\kl(amalg){strongly effective}, and so is the class of context-free languages.
Therefore, \cref{infix-wqo-is-emptiness:thm} provides a concrete decision
procedure for these classes.

\subsection{Amalgamation Systems}
\label{amalgamation-systems:subsec}

Let us now formally introduce the notion of \kl{amalgamation systems}, and
recall some results from \cite{ASZZ24} that will be useful for the proof of
\cref{infix-amalgamation:thm}. The notion of \kl{amalgamation system} is
tailored to produce \emph{pumping arguments}, which is exactly what our
\cref{pumping-periods:lem} talks about. At the core of a pumping argument,
there is a notion of \emph{run}, which could for instance be a sequence of
transitions taken in a finite state automaton. Continuing on the analogy with
finite automata, there is a natural ordering between runs, i.e., a run is
smaller than another one if one can ``unroll loops" of the first run to obtain
the second one. Typical pumping arguments then rely on the fact that
\emph{minimal} runs are of finite size, and that all other runs are
obtained by ``gluing'' minimal runs together. This is exactly what
\kl{amalgamation systems} are about.

@@

Leveraging a similar reasoning, we conclude that being a \kl{bounded language},
a \kl{regular language}, or being included in a finite union of products of
\kl{chains} all collapse for \kl{well-quasi-ordered} and \kl{downwards closed}
languages for the \kl{infix relation}.

@@

\AP Combining
\cref{thue-morse:lemma,dwclosed-infixes-wqo:lem}, we can
conclude that the collection of \kl{infixes} of the \kl{Thue-Morse sequence}
cannot be recognized by \emph{any} \kl{amalgamation system}. To put this result
in context, it was proven that the complement of the set $T$ of prefixes of the
\kl{Thue-Morse sequence} is context-free, and conjectured that the same holds
for the complement of $\LMorse$ \cite{BERST86}. To the knowledge of the
authors, this conjecture remains open. Our remark provides a negative answer to
this conjecture for the language $\LMorse$.

@@

\AP Let us now introduce our effectiveness assumptions on \kl{amalgamation
systems}. We follow the approach of \cite{ASZZ24} and require that an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$ is effective when $R$ is
recursively enumerable, the function $\canrun(\cdot)$ is computable, and for
any two runs $r, s \in R$, the set $E(r,s)$ is computable.

\todo[inline]{In this definition, the alphabet $\Sigma$ can be changed 
(and we seem to actually use it in the proofs), it is strange.}

\AP We say that a class $\mathcal{C}$ of languages is an \kl{effective
amalgamative class} whenever for every $L \in \mathcal{C}$, there exists an
effective \kl{amalgamation system} recognizing $L$, and such that $\mathcal{C}$
is \kl{effectively closed under rational transductions}. Recall that a
\intro{rational transduction} is a function $f \colon \Sigma^* \to \Gamma^*$
that can be defined by a (nondeterministic) finite automaton with output. A
class of languages $\mathcal{C}$ is \intro{effectively closed under rational
transductions} when, for every language $L \in \mathcal{C}$, and every rational
transduction $f \colon \Sigma^* \to \Gamma^*$, the image of $L$ through $f$
belongs to $\mathcal{C}$ and is effectively computable. We refer the reader to
\cite{BERST79} for a primer on \kl{rational transductions}.

\AP A class $\mathcal{C}$ of languages is \intro(amalg){strongly effective}
whenever the emptiness problem for languages in $\mathcal{C}$ is decidable.
This notion is interesting because usual language problems such as boundedness
or simultaneous boundedness are decidable for \kl{strongly effective
amalgamative classes}~\cite{ASZZ24}.

@@

    Let $R = \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in \Sigma^{\leq m
    \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup
    \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of
    the bounds $n$, $m$, and $N_0$, this language is regular. The map $L
    \mapsto L \cap \Sigma^* \setminus R$  is a \kl{rational transduction}
    because $\Sigma^* \setminus R$ is regular. Since $\mathcal{C}$ is
    \kl{closed under rational transductions}, we can therefore reduce the
    inclusion to emptiness of $L \cap \overline{R}$. However, we need to find
    these bounds first.

    To determine values for $n$ and $m$, we first test if $L$ is bounded. Since
    emptiness is decidable, we can apply the algorithm in~\cite[Section
    4.2]{ASZZ24} to decide if $L$ is bounded. If $L$ is bounded, this algorithm
    yields words $w_1, \ldots w_n$ such that $L \subseteq w_1^* \cdots w_n^*$
    and therefore yields also the bounds in questions. If $L$ is not bounded,
    then $L$ cannot be \kl{well-quasi-ordered} by the \kl{infix relation}
    because of \cref{infix-amalgamation:thm} and we immediately return false.

@@

    \cref{wqo-infix-decidable} $\Rightarrow$
    \cref{wqo-prefix-decidable}. We just consider the transduction $f$
    that maps every word $w$ to $\# w$ where $\# $ is a fresh symbol. Then, for
    any language $L \in \mathcal C$, $L$ is \kl{well-quasi-ordered} by
    \kl{prefix} if and only if $f(L)$ is \kl{well-quasi-ordered} by \kl{infix}.
    \todo{fresh symbols? can we assume we have those??}

    \cref{wqo-prefix-decidable} $\Rightarrow$
    \cref{emptiness-decidable}. We consider the transduction $f$ that 
    maps every word $w$ to the word $a b^{|w|} a$.
    Then, for any language $L \in \mathcal C$, $f(L)$ is \kl{well-quasi-ordered}
    by \kl{prefix} if and only if $L$ is finite if and only if $L$ is finite.
    Hence, we can decide if $L$ is finite, and therefore if $L$ is empty 
    by enumeration.
    \todo{I don't understand the notation $\Sigma^* \times \set{a,b}^*$, so I did 
    a worse but correct proof for me.}
\end{proofof}

### 2024-10-18 src/infixes.tex
new proofs

\begin{lemma}
    \label{thue-morse:lemma}
    There exists a language $L$ that is \kl{downwards closed} for the \kl{infix
    relation}, \kl{well-quasi-ordered} for the \kl{infix relation}, but is not
    \kl(language){bounded}.
\end{lemma}
\begin{proof}
    Let $\ThueMorse$ be the \kl{Thue-Morse sequence}
    $L$ be the set of infixes of $w$. By construction $L$ is an \emph{infinite}
    \kl{downwards closed} for the \kl{infix relation}. Let us argue that $L$ is
    \kl{well-quasi-ordered} for the \kl{infix relation}, but is not \kl(language){bounded}.

    Assume by contradiction that $L$ is \kl(language){bounded}. In this case, there exist
    words $w_1, \dots, w_k \in \Sigma^*$ such that $L \subseteq w_1^* \cdots
    w_k^*$. Since $L$ is infinite and \kl{downwards closed}, there exists a
    word $u \in L$ such that $u = w_i^3$ for some $1 \leq i \leq k$. This is absurd
    because $u \infleq \ThueMorse$, which is \kl{cube-free}.

@@

\begin{lemma}
    \label{thue-morse-ordinal:lemma}
    Let $L$ be the language of \kl{infixes} of the \kl{Thue-Morse sequence}.
    Then, the \kl{ordinal invariants} of $L$ are as follows:
    the \kl{maximal order type} of $L$ is $\omega$,
    the \kl{ordinal height} of $L$ is $\omega$,
    the \kl{ordinal width} of $L$ is $\omega$.
\end{lemma}
\begin{proof}
    Let us prove that these are upper bounds for the \kl{ordinal invariants} of
    $L$. For the \kl{ordinal height}, it is true for any language $L$.
    For the \kl{maximal order type}, we remark that
    the maximal length of a \kl{bad sequence} is determined by its first element,
    hence that it is at most $\omega$.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order type},
    we conclude that the \kl{ordinal width} is also at most $\omega$.

    Now, let us prove that these bounds are tight. It is clear that
    $\oHeight{L} = \omega$: given any number $n \in \Nat$, one can construct a
    \kl{decreasing sequence} of words in $L$ of length $n$, for instance by
    considering the first $n$ prefixes of the \kl{Thue-Morse sequence} by
    decreasing size.
    Let us now prove that $\oWidth{L} = \omega$. Assume by contradiction that
    $\oWidth{L}$ is finite. Then, $L$ can be written as a finite union of
    \kl{chains} for the \kl{infix relation}, and in particular, $L$ is
    \kl(language){bounded}, which is absurd by \cref{thue-morse:lemma}.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order
    type}, we conclude that the \kl{maximal order type} of $L$ is also $\omega$.
\end{proof}

>>>>

\begin{lemma}
    \label{thue-morse:lemma}
    The language $\intro*\LMorse$ of \kl{infixes} of the \kl{Thue-Morse sequence}
    is \kl{downwards closed} for the \kl{infix
    relation}, \kl{well-quasi-ordered} for the \kl{infix relation}, but is not
    \kl(language){bounded}.
\end{lemma}
\begin{proof}
    By construction $\LMorse$ is an \emph{infinite}
    \kl{downwards closed} for the \kl{infix relation}. Let us argue that $\LMorse$ is
    \kl{well-quasi-ordered} for the \kl{infix relation}, but is not \kl(language){bounded}.

    Assume by contradiction that $\LMorse$ is \kl(language){bounded}. In this case, there exist
    words $w_1, \dots, w_k \in \Sigma^*$ such that $L \subseteq w_1^* \cdots
    w_k^*$. Since $L$ is infinite and \kl{downwards closed}, there exists a
    word $u \in L$ such that $u = w_i^3$ for some $1 \leq i \leq k$. This is absurd
    because $u \infleq \ThueMorse$, which is \kl{cube-free}.

@@

\begin{lemma}
    \label{thue-morse-ordinal:lemma}
    The \kl{maximal order type} of $\LMorse$ is $\omega$,
    the \kl{ordinal height} of $\LMorse$ is $\omega$,
    the \kl{ordinal width} of $\LMorse$ is $\omega$.
\end{lemma}
\begin{proof}
    Let us prove that these are upper bounds for the \kl{ordinal invariants} of
    $L$. For the \kl{ordinal height}, it is true for any language $\LMorse$.
    For the \kl{maximal order type}, we remark that
    the maximal length of a \kl{bad sequence} is determined by its first element,
    hence that it is at most $\omega$.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order type},
    we conclude that the \kl{ordinal width} is also at most $\omega$.

    Now, let us prove that these bounds are tight. It is clear that
    $\oHeight{\LMorse} = \omega$: given any number $n \in \Nat$, one can construct a
    \kl{decreasing sequence} of words in $L$ of length $n$, for instance by
    considering the first $n$ prefixes of the \kl{Thue-Morse sequence} by
    decreasing size.
    Let us now prove that $\oWidth{\LMorse} = \omega$. Assume by contradiction that
    $\oWidth{\LMorse}$ is finite. Then, $L$ can be written as a finite union of
    \kl{chains} for the \kl{infix relation}, and in particular, $L$ is
    \kl(language){bounded}, which is absurd by \cref{thue-morse:lemma}.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order
    type}, we conclude that the \kl{maximal order type} of $\LMorse$ is also $\omega$.
\end{proof}
